Attributions for Poverty in Post-Socialist Countries

### 2014-02-19 article2013.Rmd
Mikon muutokset!

The purpose of this article is to examine cross-country differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and Eastern European countries (Gallie and Paugam, 2002; Kallio and Niemelä, 2013; Lepianka, 2007; van Oorschot and Halman, 2000). East Europeans are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in East Europeans perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the East and the West and within post-socialist regime as well as on the importance of the post-socialist legacy. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crises both at the individual and contextual level and their association with attributions for poverty. At the individual level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

The article is structured as follows: A brief overview of theoretical discourse on explanations of poverty and prior findings on cross-national differences in and contextual level determinants of attributions for poverty is presented in Section 2. Section 3 focuses on the self-interest approach as an individual level explanatory factor of the perceptions with particular interest of risk factors related to individual’s financial situation.  Research design, survey data and methods are described in detail in Section 4. The results of the study are presented in Section 5. Finally, the main conclusions, implications and limitations of the study are discussed in Section 6.

@@

National case studies from Estonia and Russia have given support to findings of cross-national studies. There is strong support for the idea that the poor have only themselves to blame for their poverty. In addition, both of these countries have strong support for social blame as well. Estonians are as likely to blame the flaws of the economic system as the individualistic reasons. In Russia, on the other hand, the economic system is blamed even more than the individuals themselves (Gorshkov and Tikhonova, 2006; Stephenson, 2000). These results are consistent with Lepianka's (2007: 127–135) comparative analysis of Hungary, Russia, Czech Republic and Estonia. Interesting result is that social blame coincides strongly with individual blame. This illustrates the split-consciousness of public perceptions, which means the coexistence of both dominant and potentially challenging beliefs, where the former represents dominant stratification ideology or value structure of a given country and the latter everyday stratification-related experience of an individual (Kluegel et al., 1995).

Based on qualitative poverty studies in Central Asia, Caucasus and Ukraine and Moldova Dudwick et al. (2003: 23) states that, generally, people blamed their own poverty on the failure of the Soviet state and the corruption, indifference, and incompetence of their new leaders. At the same time, however, they often attributed the poverty of others to individual failure, such as laziness, alcoholism, having too many children to support, or having too few children to provide for their old age. Emphazising the positive association between horizontal and vertical trust (see e.g. Rothstein, 2000), they also reported that many of the respondents had become very distrustful of government officials and many suspected that government officials of exploiting the political and economic turmoil for their own gain, at the expense of ordinary citizens. With extensive country selection, the data utilized in this study provides a marvellous possibility to examine the association between attributions for poverty and the context regarding political process, civil liberties and political rights. We can assume that in those countries where the state of democracy is low, citizens blame more likely society for poverty than in those countries where citizens have political freedom and power to influence socio-political issues.

Besides the so called good government, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to East-West differences as well as variation within post-socialist countries, countries examined in this study have faced global economic crises in distinct ways. While many Western and Eastern European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crises, the research design of this article provides empirical evidence on the effects of crises to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune, 2007). In regard to explanations of poverty, Gallie and Paugam (2002:21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2013). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

In addition to macro level economic performance, the level overall of well-being and living conditions are found to be linked with poverty attributions. Lepianka et al. (2010:65,86) found out that people living in more developed countries they tend to attribute living in destitution rather to individual unluckiness than to modern progress, but they found no link between development and either type of blame explanations. In contrast to previous studies, variation in human development within our selection of countries if far greater. Therefore we are able to compare poor countries of Central Asia with well-off Slovenia and Czech Republic. Given the great differences in living conditions within the post-socialist space, we assume that in countries of low human development and fewer welfare institutions people tend to attribute poverty more with social reasons than individual.

@@

In regard to individual-level determinants, popular attitudes toward the welfare state have traditionally been assumed to be dependent on long-term class-related interest, short-term self-interest and the values and norms that have been internalized by the individuals in the society in question (Taylor-Gooby, 1985; Svallfors, 1995). More recent directions of comparative literature on welfare attitudes tend to also focus on the distribution of marketable capacities such as skill specificity (Iversen and Soskice, 2001) and resource/class-based risks (Cusack et al. 2006; Blomberg et al. 2012; Finseraas and Ringdal, 2012) as key to welfare attitudes. Thus, studies have assumed – and to some extent proved – that groups that have a greater risk of facing social problems or being dependent on welfare state benefits might perceive the government’s role in welfare issues in a distinct way compared to those who are not exposed to these risks.

Inspired by the studies which emphasise risks as the key mechanisms, this article tests to what extent factors related to financial hardship are associated – beyond the welfare regime – with perceptions of the causes of poverty. Following this line of reasoning, we are interested in whether those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Thus, the research design in regard to determinants of poverty attributions enables us to study from a fairly unexplored perspective the different mechanisms through which individuals who have a greater risk to experience material destitution perceive the causes of poverty.

@@

Not only the current situation or the perceived future but also the past experiences might be associated with individual’s evaluations (e.g. Kumlin, 2007). In regard to attributions for poverty, the effects of past experiences are – according to our knowledge – totally unexplored area of research. In this article we will analyse past experiences by two ways. First, we are interested in whether those whose current economic situation is lower than four years ago perceive the causes of poverty distinct way compared to those whose economic situation has improved or remained in the same level. Second, we will focus on the effects of the global financial crisis. We assume that economic crisis have affected especially those who are the most vulnerable groups in the society. Therefore, we will analyse whether those who report that the economic crisis has affected their household at least a fair amount perceive the causes of poverty differently than those who report that it has had no or only a little effect to their household. According to self-interest hypothesis, we assume that those whose economic situation has declined and those whose household has affected by economic crisis are more inclined to blame society for poverty than those who has not experienced economic problems.

@@

Thus, the question is a standard forced-choice question which is used also in other comparative surveys such as in World Values Survey and in Eurobarometer. According to van Oorschot and Halman (2000) different statements represent a four-tier typology of poverty explanations as follows: While the unlucky refers to individual fate, the laziness and lack of willpower represents individual blame. In addition, injustice in a society indicates social blame, whereas the view that poverty is simply part of modern life refers to social fate (see figure 1). Based on the findings of previous studies, we assume that in general post-socialist countries endorse more likely individual blame and social blame types of explanations than individual and social fate types of explanations. However, we hypothesize also that findings indicate relatively large variation between countries.

We target our analyses at 22 post-socialist countries and use five Western European countries as references in descriptive analyses (see figure 1.). Our selection on 22 post-socialist countries includes Czech Republic, Estonia, Hungary, Bulgaria, Latvia, Lithuania, Poland, Slovakia, Slovenia, Romania from Central Eastern Europe (CEE), and Armenia, Azerbaijan, Belarus, Georgia, Kazakhstan, Kyrgyzstan, Moldova, Tajikistan, Ukraine, Uzbekistan and Russia from Commonwealth of Independent States (CIS)[^***]. Of Western European countries we have included France, Germany, Italy, United Kingdom and Sweden as references.

@@

As contextual level predictors we use country’s economic performance during the period of global financial crises, living conditions and the state of democracy. As described in previous sections, we assume that in those countries where the economic crises has meant declining economic conditions, public are more likely to endorse social explanations for poverty than in countries where national economy has continued to grow despite the global economic turbulence. Based on the theoretical discussion on civil liberties, political rights and institutional trust, we assume that the public support to social blame type of explanation is high in those countries where the state of democracy is low. The case with human deveplopment in less clear, but we assume, against the hypothesis by Lepianka et al. (2010), that lower level of human development and lack of welfare institutions may well be connected with higher support for individual blame type of explanation.

As an individual-level determinants for the perceptions article test to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty, and whether these factors are associated with perceptions similar way regardless of the  groups of countries studied. Hereby, the research design aims at contributing to the theoretical reasoning regarding the importance of self-interest for attributions for poverty which assumes that those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Analyses include six individual-level predictors, namely transfer dependency, low education, perceived low income level, income-level compared to the past and expected change in income in the future as well as perceived effect of the financial crises. Variables are recoded as binomial where value 1 is coded for risk and 0 for absence of risk. Both contextual and individual-level variables are described in the Table 1.

@@

Effect of financial crises        Whether respondents perceives that their household has suffered the financial crisis great or fair amount                             lits

@@

An additional dataset of selected macro indicators were added to the LiTS2-dataset consisting of three variables: 1) change in total GDP in 2007 - 2010, 2) Human Deveplopment Index in 2010 and 3) Voice and accountability indicator. Data for change in GDP is from World Bank's [World Development Indicators](http://data.worldbank.org/data-catalog/world-development-indicators), Human Development Index from [United Nations Development Programme](http://hdr.undp.org/en/data) and Voice and Accountability from the World Bank's [Worldwide Governance Indicators](http://info.worldbank.org/governance/wgi/index.aspx#home). All macro level indicators are obtained from [Quality of Government Institutes](http://www.qog.pol.gu.se/) data using [rQog](http://markuskainu.fi/rqog/)-package (Kainu 2014).

The empirical section starts by presenting the results of descriptive analyses. As we assume that individual level perceptions are dependent both on individual and country-level contextual factors and we have a hierarchical two level dataset, it is recommended to use multilevel regression analysis for this task (Snijders & Bosker, 2011) as multilevel regression analysis allows the estimation of country-level effects on individual level outcomes. The analysis is performed using glmer-function from lme4 package in R. All of the models employed are so called multilevel logistic random intercept models at the level of individuals and countries. Model summary tables include both a fixed effects and a random effects part. The fixed effects part will include odds ratios and statistical significances of independent variables, while the random effects part will indicate the country-level variance and standard deviation, deviance and number of cases and groups.

We are modelling the date in stepwise manner first running the empty models for both social blame (table 3) and individual blame (table 4) type of explanation. At second stage we include all individual level predictors and in third stage we include only the macro level predictors. In final models we include both the individual and macro level variables. As we are using *logistics regression* instead of *multinomial logistic regression* with dependent variable of nominal scale, we confirm our findings by running a further model where we predict the support for social blame against the support for individual blame using both the individual and contextual level variables. This model is compared with full models from tables 3 and 4 in table 5.

@@

In regard to differences between the East and the West of Europe, results show that there are some differences. Yet there are also substantial variation within Western Europe, which correspond earlier findings on cross-national differences in Western Europe (Lepianka et al., 2010; Kallio and Niemelä, 2010). Clearest differences between the West and the East is the stronger support for social fate in Western Europe. In Germany and Sweden external reasons for poverty receive the majority of support. On the other hand, attributions for poverty in France and Italy seems to be quite similar with post-socialist countries - blame types of explanations are endorsed more likely than fate types of explanations. Finally, the figures of Great Britain are the example of selective welfare policy dominated liberal regime in which the individual blame type of explanations is strongly supported (see also Albrekt Larsen, 2006).

Hence, the results are in line with our assumptions and prior research (Lepianka et al., 2010) regarding the general support for different explanations of poverty. Overall, the results indicate that for the most of the post-socialist countries two most popular explanations are individual and social blame types of explanations.Yet, the differences between social fate and individual blame types of explanations are in some countries very small. In general, Western Europeans (except Britons) do not support individual blame type of explanation as much as citizens in the East. Instead they emphasise more social reasons for poverty. Results also clearly emphasise that the idea that poverty is caused by just individual bad luck is the least popular perception of the reason for poverty.

>>>>

The purpose of this article is to examine cross-national differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and Eastern European countries (Gallie and Paugam, 2002; Kallio and Niemelä, 2013; Lepianka, 2007; van Oorschot and Halman, 2000). East Europeans are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in East Europeans' perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the East and the West and within post-socialist regime as well as on the importance of the post-socialist legacy. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crisis both at the individual and contextual level and their association with attributions for poverty. At the individual level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

The article is structured as follows: A brief overview of theoretical discourse on explanations of poverty and prior findings on cross-national differences in and contextual level determinants of attributions for poverty is presented in Section 2. Section 3 focuses on the self-interest approach as an individual-level explanatory factor of the perceptions with particular interest of risk factors related to individual’s financial situation.  Research design, survey data and methods are described in detail in Section 4. The results of the study are presented in Section 5. Finally, the main conclusions, implications and limitations of the study are discussed in Section 6.

@@

National case studies from Estonia and Russia have given support to findings of cross-national studies. There is strong support for the idea that the poor have only themselves to blame for their poverty. In addition, both of these countries have strong support for social blame as well. Estonians are as likely to blame the flaws of the economic system as the individualistic reasons. In Russia, on the other hand, the economic system is blamed even more than the individuals themselves (Gorshkov and Tikhonova, 2006; Stephenson, 2000). These results are consistent with Lepianka's (2007: 127–135) comparative analysis of Hungary, Russia, Czech Republic and Estonia as well as with Habibov's (2011) comparative evidence from the Caucasus, Central Asia, Russia and Ukraine. Interesting result is that social blame coincides strongly with individual blame. This illustrates the split-consciousness of public perceptions, which means the coexistence of both dominant and potentially challenging beliefs, where the former represents dominant stratification ideology or value structure of a given country and the latter everyday stratification-related experience of an individual (Kluegel et al., 1995).

Based on qualitative poverty studies in Central Asia, the Caucasus, Ukraine and Moldova, Dudwick et al. (2003: 23) states that, generally, people blamed their own poverty on the failure of the Soviet state and the corruption, indifference, and incompetence of their new leaders. At the same time, however, they often attributed the poverty of others to individual failure, such as laziness, alcoholism, having too many children to support, or having too few children to provide for their old age. Emphazising the positive association between horizontal and vertical trust (see e.g. Rothstein, 2000), they also reported that many of the respondents had become very distrustful of government officials and many suspected that government officials of exploiting the political and economic turmoil for their own gain, at the expense of ordinary citizens. With extensive country selection, the data utilized in this study provides a marvellous possibility to examine the association between attributions for poverty and the context regarding political process, civil liberties and political rights. We can assume that in those countries where the state of democracy is low, citizens blame more likely society for poverty than in those countries where citizens have political freedom and power to influence socio-political issues.

Besides the so called good government, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to East-West differences as well as variation within post-socialist countries, countries examined in this study have faced global economic crisis in distinct ways. While many Western and Eastern European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crisis, the research design of this article provides empirical evidence on the effects of crisis to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune, 2007). In regard to explanations of poverty, Gallie and Paugam (2002: 21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2013). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

In addition to macro-level economic performance, living conditions in general are found to be linked with attributions for poverty. People living in countries with high unemployment are more likely to endorse social than individualistic explanations (Albrekt Larsen, 2006: 74-79). In addition, the increased incidence of poverty reduces the likelihood of choosing individual laziness as an explanation of poverty (Lepianka, 2007: 118-121). However, there are to some extent contradictory findings as well. While Kallio and Niemelä (2013) did not find association between individual blame type of explanation and incidence of poverty, Lepianka et al. (2010) found out that people living in more developed countries tend to attribute poverty rather to individual and social fate than blame types of explanations. Yet, in contrast to previous studies, the variation between post-socialist countries is far larger than in Western Europe. Consequently, by comparing for instance poorer countries of Central Asia with more well-off Slovenia and Czech Republic, this study provides a possibility to analyze whether the association between attributions for poverty and living conditions is also stronger in the post-socialist countries. This study examines human development index (HDI) as a proxy for living conditions in a given country. Given the great differences in living conditions within the post-socialist space, we assume that in countries of low level of HDI people tend to attribute poverty more with social than individual reasons.

@@

In regard to individual-level determinants, popular attitudes toward the welfare state have traditionally been assumed to be dependent on long-term class-related interest, short-term self-interest and the values and norms that have been internalized by the individuals in the society in question (Taylor-Gooby, 1985; Svallfors, 1995). More recent directions of comparative literature on welfare attitudes tend to also focus on the distribution of marketable capacities such as skill specificity (Iversen and Soskice, 2001) and resource/class-based risks (Cusack et al., 2006; Blomberg et al., 2012; Finseraas and Ringdal, 2012) as key to welfare attitudes. Thus, studies have assumed – and to some extent proved – that groups that have a greater risk of facing social problems or being dependent on welfare state benefits might perceive the government’s role in welfare issues in a distinct way compared to those who are not exposed to these risks.

Inspired by the studies which emphasize risks as the key mechanisms, this article tests to what extent factors related to financial hardship are associated with perceptions of the causes of poverty. Following this line of reasoning, we are interested in whether those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Thus, the research design in regard to determinants of poverty attributions enables us to study from a fairly unexplored perspective to the different mechanisms through which individuals who have a greater risk to experience material destitution perceive the causes of poverty.

@@

Not only the current situation or the perceived future but also the past experiences might be associated with individual’s evaluations (e.g. Kumlin, 2007). In regard to attributions for poverty, the effects of past experiences are – according to our knowledge – totally unexplored area of research. In this article we will analyse past experiences by two ways. First, we are interested in whether those whose current economic situation is lower than four years ago perceive the causes of poverty distinct way compared to those whose economic situation has improved or remained in the same level. Second, we will focus on the effects of the global financial crisis. We assume that economic crisis have affected especially those who are the most vulnerable groups in the society. Therefore, we will analyze whether those who report that the economic crisis has affected their household at least a fair amount perceive the causes of poverty differently than those who report that it has had no or only a little effect to their household. According to self-interest hypothesis, we assume that those whose economic situation has declined and those whose household has affected by economic crisis are more inclined to blame society for poverty than those who has not experienced economic problems.

@@

Thus, the question is a standard forced-choice question which is used also in other comparative surveys such as in World Values Survey and in Eurobarometer. According to van Oorschot and Halman (2000) different statements represent a four-tier typology of poverty explanations as follows: While the unlucky refers to individual fate, the laziness and lack of willpower represents individual blame. In addition, injustice in a society indicates social blame, whereas the view that poverty is simply part of modern life refers to social fate (see Figure 1). Based on the findings of previous studies, we assume that in general post-socialist countries endorse more likely individual blame and social blame types of explanations than individual and social fate types of explanations. However, we hypothesize also that findings indicate relatively large variation between countries.

We target our analyses at 22 post-socialist countries and use five Western European countries as references in descriptive analyses (see Figure 2). Our selection on 22 post-socialist countries includes Czech Republic, Estonia, Hungary, Bulgaria, Latvia, Lithuania, Poland, Slovakia, Slovenia, Romania from Central Eastern Europe (CEE), and Armenia, Azerbaijan, Belarus, Georgia, Kazakhstan, Kyrgyzstan, Moldova, Tajikistan, Ukraine, Uzbekistan and Russia from Commonwealth of Independent States (CIS)[^***]. Of Western European countries we have included France, Germany, Italy, United Kingdom and Sweden as references.

@@

As contextual level predictors we use country’s economic performance during the period of global financial crisis, living conditions and the state of democracy. As described in previous sections, we assume that in those countries where the economic crisis has meant declining economic conditions, public are more likely to endorse social explanations for poverty than in countries where national economy has continued to grow despite the global economic turbulence. Based on the theoretical discussion on civil liberties, political rights and institutional trust, we assume that the public support to social blame type of explanation is high in those countries where the state of democracy is low. The case with human deveplopment is to some extent less clear, but we assume that the lower level of human development is connected with stronger support for social than individual explanation.

As an individual-level determinants for the perceptions article test to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty, and whether these factors are associated with perceptions similar way regardless of the  groups of countries studied. Hereby, the research design aims at contributing to the theoretical reasoning regarding the importance of self-interest for attributions for poverty which assumes that those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Analyses include six individual-level predictors, namely transfer dependency, low education, perceived low income level, income-level compared to the past and expected change in income in the future as well as perceived effect of the financial crisis. Variables are recoded as binomial where value 1 is coded for risk and 0 for absence of risk. Both contextual and individual-level variables are described in the Table 1.

@@

Effect of financial criris        Whether respondents perceives that their household has suffered the financial crisis great or fair amount                             lits

@@

An additional dataset of selected macro indicators were added to the LiTS2-dataset consisting of three variables: 1) change in total GDP in 2007 - 2010, 2) Human Deveplopment Index in 2010 and 3) Voice and accountability indicator. Data for change in GDP is from World Bank's [World Development Indicators](http://data.worldbank.org/data-catalog/world-development-indicators), Human Development Index from [United Nations Development Programme](http://hdr.undp.org/en/data) and Voice and Accountability from the World Bank's [Worldwide Governance Indicators](http://info.worldbank.org/governance/wgi/index.aspx#home). All macro level indicators are obtained from [Quality of Government Institutes](http://www.qog.pol.gu.se/) data using [rQog](http://markuskainu.fi/rqog/)-package (Kainu, 2014).

The empirical section starts by presenting the results of descriptive analyses. As we assume that individual level perceptions are dependent both on individual and country-level contextual factors and we have a hierarchical two level dataset, it is recommended to use multilevel regression analysis for this task (Snijders and Bosker, 2011) as multilevel regression analysis allows the estimation of country-level effects on individual level outcomes. The analysis is performed using glmer-function from lme4 package in R. All of the models employed are so called multilevel logistic random intercept models at the level of individuals and countries. Model summary tables include both a fixed effects and a random effects part. The fixed effects part will include odds ratios and statistical significances of independent variables, while the random effects part will indicate the country-level variance and standard deviation, deviance and number of cases and groups.

We are modelling the date in stepwise manner first running the empty models for both social blame (Table 3) and individual blame (Table 4) type of explanation. At second stage we include all individual-level predictors and in third stage we include only the macro-level predictors. In final models we include both the individual and macro-level variables. As we are using *logistics regression* instead of *multinomial logistic regression* with dependent variable of nominal scale, we confirm our findings by running a further model where we predict the support for social blame against the support for individual blame using both the individual and contextual level variables. This model is compared with full models from tables 3 and 4 in Table 5.

@@

In regard to differences between the East and the West of Europe, results show that there are some differences. Yet, there are also substantial variation within Western Europe, which correspond earlier findings on cross-national differences in Western Europe (Lepianka et al., 2010; Kallio and Niemelä, 2010). Clearest differences between the West and the East is the stronger support for social fate in Western Europe. In Germany and Sweden external reasons for poverty receive the majority of support. On the other hand, attributions for poverty in France and Italy seems to be quite similar with post-socialist countries - blame types of explanations are endorsed more likely than fate types of explanations. Finally, the figures of Great Britain are the example of selective welfare policy dominated liberal regime in which the individual blame type of explanations is strongly supported (see also Albrekt Larsen, 2006).

Hence, the results are in line with our assumptions and prior research (Habibov, 2011; Lepianka et al., 2010) regarding the general support for different explanations of poverty. Overall, the results indicate that for the most of the post-socialist countries two most popular explanations are individual and social blame types of explanations.Yet, the differences between social fate and individual blame types of explanations are in some countries very small. In general, Western Europeans (except Britons) do not support individual blame type of explanation as much as citizens in the East. Instead they emphasize more social reasons for poverty. Results also clearly indicate that the idea that poverty is caused by just individual bad luck is the least popular perception of the reason for poverty.
